A Note on Convergence Diagnostics for Multiple Imputation Algorithms

### 2019-12-15 Simulation/Writeup/NoteConvergence.tex
Presentation added

%The potential scale reduction factor $\widehat{R}$ is obtained as:
We can interpret $\widehat{R}$ as potential scale reduction factor since it indicates by how much the variance of $\theta$ could be shrunken down if an infinite number of iterations per chain would be run \citep{gelm92}. This interpretation assumes that chains are `over-dispersed' at $t=1$, and reach convergence as $T \to \infty$. Over-dispersion implies that the initial values of the chains are `far away' from the target distribution and each other. The iteration at which chains are independent of their initial values is the point where mixing occurs. This is diagnosed with $\widehat{R}$-values close to one. %If there is no added value 
High $\widehat{R}$-values thus indicate slow mixing of chains, which is a sign of non-convergence. %, e.g., because chains may depend on their initial value. %are of dependence across chains on initial values. Once there is no dependence on initial values anymore, there is no added value of continuing iterations. \textbf{[MOGELIJK OVERBODIGE ZIN HIER]} The measure then diagnoses algorithmic convergence ($\widehat{R} = 1$). 
%If there is no added value compared to infinitely many iterations, the chains have converged. 
The conventionally acceptable threshold for this convergence diagnostic was $\widehat{R} < 1.2$ \citep{gelm92}. Recently, \cite{veht19} proposed a more stringent threshold: $\widehat{R} < 1.01$.

@@

We can interpret %the sign and magnitude of ... Any non-zero $AC$-value ($-1 \leq AC \leq 1$) . $AC = 0$ indicates independence between $\theta$-values.
$AC$-values as a measure of dependence within chains. Positive $AC$-values indicate recurrence (i.e., similarity between $\theta$-values of subsequent iterations), which may lead to trending. Negative $AC$-values occur when the $\theta$-values of subsequent iterations diverge from one-another. Dependence in this sense does not threaten the stationarity component of convergence. %that we aim to evaluate with $AC$. 
On the contrary even---it may increase the variance of $\theta$. \textbf{speed up convergence???}  %have also indicate dependence, but do not threaten the stationarity component of convergence \textbf{[REMOVE INTERPRETATION HERE?]}. Instead of recurrence, there would be a `seesaw' effect within chains: subsequent $\theta$-values would deviate in opposite directions from the chain average $\theta^{(\cdot m)})$, increasing the variance of $\theta$. 
As convergence diagnostic, the interest is in positive $AC$-values.%\footnotemark 
\footnote{The magnitude of $AC$-values can also be evaluated statistically, but that is outside of this note's scope.}

@@

This study evaluates whether $\widehat{R}$ and $AC$ could diagnose convergence of multiple imputation algorithms. There is, however, no baseline measure available to evaluate their performance against.
\footnote{The current practice of visually inspecting imputation chains for signs of non-mixing or non-stationarity can only diagnose severely pathological cases of non-convergence.} 
We will assess the appropriateness of the two convergence diagnostics in comparison to evaluation criteria recommended by \cite[\S~2.5.2]{buur18}. The performance diagnostics comprise of average bias, average confidence interval width, and empirical coverage rate across simulations.% %"We expect that in accordance with Neyman (1934), at least 95 percent of the confidence intervals should contain the true population value, although some room for simulation error should be taken into account to counteract the finite nature of our simulations"\cite{schou18}.
\footnote{We could also look at distributional characteristics, and plausibility of imputed values, see \cite{vinknd} (n.d.). For now, this is outside of the scope of this study.} % Or use: Then evaluate the regular diagnostics of MI simulations, see Vink, n.d., and  convergence. Diagnostics include absolute bias of the estimated regression coefficient, confidence interval width, and empirical coverage rate across simulations. Convergence is evaluated with $\widehat{R}$, or potential scale reduction factor.

\textbf{[HIER NOG ZOOI]} We hypothesize that $\widehat{R}$ may not be an appropriate diagnostic, because it assumes over-dispersed initial values. In \pkg{mice}, initial values of the algorithm are chosen randomly from the observed data. Therefore, we cannot be certain that the initial values are over-dispersed.
%The initial parameter value is noodzakelijk. Daarom is het lastig om in MI that starten vanuit overdisp state. 
% Starting values of MICE chains: "For each j, fill in starting imputations $Y^0_j$ by random draws from $Y^{obs}_j$." \cite[\S~4.5.2]{buur18} before for loops over variables and imputations, see \url{https://stefvanbuuren.name/fimd/sec-FCS.html#def:mice}
Without over-dispersed initial states, $\widehat{R}$ may falsely diagnose convergence \citep{broo98}. %would not be sensitive enough to flag non-convergence of MI algorithms: "if over-dispersion does not hold, $\sigma_+^2$ can be too low, which can lead to falsely diagnosing convergence." \cite[p~437]{broo98}. 
This suggests that $\widehat{R}$ would not be sensitive enough to flag non-convergence of MI algorithms. However, an empirical finding by \cite{lace07} shows that $\widehat{R}$ can indicate non-convergence at unconventionally long chain lengths. While the default number of iterations in \pkg{mice} is five, $\widehat{R} > 1.1$ was observed after 50 iterations, and $\widehat{R} > 1.01$ up-to 150 iterations. This suggests that $\widehat{R}$ would over-estimate non-convergence of MI algorithms.

\textbf{We hypothesize that $\widehat{R}$ will over-estimate non-convergence. No hypothesis was formulated about AC, results are exploratory.}
%
%

@@

%To investigate the convergence properties of the multiple imputation algorithm in \pkg{mice}, w
We use model-based simulation in \proglang{R} to perform multiple imputation under 100 simulation conditions. In each condition, the MI algorithm comprises a different imputation chain length ($T = 1, 2, \dots, 100$). The number of simulation runs per condition is 1000. For each simulation condition (i.e., number of iterations) and each repetition (i.e., simulation run) we compute convergence diagnostics ($\widehat{R}$ and autocorrelation) and simulation diagnostics (bias, confidence interval width and coverage). These diagnostics are aggregated across repetitions to obtain averages per simulation condition.
%We compute $\widehat{R}$ and autocorrelation (`convergence diagnostics') and performance measures (`simulation diagnostics') for each condition. %All simulation conditions are replicated 1000 times. %For each of the 100 conditions 1000 simulation runs are performed. for each of the 100 simulation conditions. \textbf{Estimands are convergence diagnostics. Performance measures are simulation diagnostics.  where for one through one hundred the Rhat and ... are calculated.} 
The simulation set-up is summarized in the pseudo-code below.\footnote{The complete \proglang{R} script of the simulation study is available from \href{https://github.com/gerkovink/shinyMice/simulation}{github.com/gerkovink/ShinyMICE/simulation}.}

@@

\subsection{Amputation}
The complete data is `amputed' once in each simulation repetition. That is, the \pkg{mice} function \fct{ampute} is used to impose a missingness mechanism upon the data. The missingness is univariate, and the probability to be missing is the same for all variables, namely 20\% (\texttt{prop = 0.8, mech = "MCAR"}). This leaves 20\% of the rows completely observed. %The resulting amputed data is equal for all simulation conditions in the same repetition. %"Furthermore, we generate three kinds of missingness proportions: 0.1, 0.5, and 0.9. Note that these values indicate the sampled proportion of incomplete cases in Y" \cite{schou18}.

\subsection{Imputation}
Missing datapoints are imputed with the \pkg{mice} function \fct{mice}. All simulations are performed with Bayesian linear regression imputation (\texttt{method = "norm"}), and five imputation chains (\texttt{m = 5}). The number of iterations varies between simulation conditions (\texttt{maxit = 1, 2, \dots, 100}). 
% Missingness mechanism: "With MCAR missingness mechanisms, the probability to be missing is the same for all cases. This is a necessary simulation condition for evaluating the performance of imputation procedures. If an imputation method is not able to solve the problem (i.e. yield valid inference) under MCAR, the statistical properties of the procedure are not sound." (Vink, n.d., p. 4)
% "combine the m completed data inferences into a single inference following Rubin's (1987) rules (pp. 76-77)" \cite{schou18}.

\subsection{Convergence Diagnostics}
% Add ``post-processing steps (e.g. aggregation computations, summary statistics, regressions on the simulation results) used to transform simulation outputs to reported results''. %, see \url{http://jsterman.scripts.mit.edu/docs/Rahmandad-Sterman%20Simulation%20Reporting%20Standards3.30%20update.pdf}.
We extract not only the imputed values of each MI procedure (where $t = T$), but also information from the intermediate iterations ($t = 1, 2, \dots, T$). The quantities of interest ($\theta$-values) are chain means and chain variances. We compute convergence diagnostics separately for each of the four variables in the dataset. \textbf{We report the maximum (absolute) value as the `worst linear predictor' of convergence [Rephrase or footnote??]}. $\widehat{R}$ is computed by implementing Vehtari et al.'s \citeyear{veht19} recommendations. $AC$ is computed with \pkg{stats} function \fct{acf}.
%%%%%%%%%%%%%%%%
%From each imputation chain, we extract imputed values, chain means and chain variances ($\theta$-values). $\theta$-values are information from the intermediate iterations ($t = 1, 2, \dots, T$). The quantities of interest ($\theta$-values) are . We compute convergence diagnostics separately for each of the four variables in the dataset. \textbf{We report the maximum (absolute) value as the `worst linear predictor' of convergence [Rephrase or footnote??]}. $\widehat{R}$ is computed by implementing Vehtari et al.'s \citeyear{veht19} recommendations. $AC$ is computed with \pkg{stats} function \fct{acf}.
%%%%%%%%%%%%%
% 
% 
% extract chain means and chain variances per iteration $t$. From the MI algorithm Convergence diagnostics are computed for chain means and chain variances.   
% $\widehat{R}$ is computed within imputation chains: . The maximum value across variables within the same simulation is reported. 
% 
% autocorrelation (AC) is computed within imputation chains,  The AC of the variable with the highest absolute AC value is reported per simulation.

@@

To estimate the quantity of scientific interest, $Q$, we perform multiple linear regression on each completed dataset with the \pkg{stats} function \fct{lm}. We obtain an estimated regression coefficient per imputation, which are pooled into a single estimate, $\bar{Q}$. We use the \pkg{mice} function \fct{pool} to get variance estimates according to Rubin's \citeyear{rubin87} rules, and utilize these to implement finite population pooling as per \citep{vink14}.

@@

We compute bias as the difference between $\bar{Q}$ and $Q$. Confidence interval width (CIW) is defined as the difference between the lower and upper bound of the 95\% confidence interval around $\bar{Q}$ (CI95\%). We compute the CI95\% bounds as

$$\bar{Q} \pm t_{(m-1)} \times SE_{\bar{Q}}, \text{ where } t_{(m-1)} = 2.66 \text{ if } \alpha = .05. $$

$t_{(m-1)}$ is the quantile of a $t$-distribution with $m-1$ degrees of freedom, and $SE_{\bar{Q}}$ is the square root of the pooled variance estimate. From bias and CIW, we calculate empirical coverage rates. Coverage rate is the proportion of simulations in which $Q$ is between the bounds of the CI95\% around $\bar{Q}$.

@@

\begin{figure}[h]
  \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
        \includegraphics{Figures/conv_diag_SEs.pdf}
  }
  \caption{Convergence diagnostics over 1000 MCMC simulations.}
    \label{fig:conv2}
\end{figure}

\begin{figure}[h]
  \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
        \includegraphics{Figures/simulation_diag.pdf}
  }
  \caption{Simulation diagnostics over 1000 MCMC simulations.}
    \label{fig:sim}
\end{figure}

\begin{figure}[h]
  \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
        \includegraphics{Figures/sim_diag_SEs.pdf}
  }
  \caption{Simulation diagnostics over 1000 MCMC simulations.}
    \label{fig:sim2}
\end{figure}

% latex table generated in R 3.6.1 by xtable 1.8-4 package
% Wed Dec 04 18:16:11 2019
\begin{table}[ht]
\centering
\caption{Simulation and convergence diagnostics over 1000 MCMC simulations.} 
\label{tab:results}
\begin{tabular}{lrrrrrrr}
  \hline
T & Bias & CI width & Cov. rate & $\widehat{R}_{mean}$ & $\widehat{R}_{var}$ & $AC_{mean}$ & $AC_{var}$ \\ 
  \hline
   1 & -0.137 & 0.954 & 0.932 & NA & NA & NA & NA \\ 
     2 & -0.006 & 0.932 & 0.953 & 1.650 & 1.632 & -0.500 & -0.500 \\ 
     3 & 0.002 & 0.929 & 0.944 & 1.314 & 1.306 & -0.660 & -0.659 \\ 
     4 & 0.003 & 0.933 & 0.957 & 1.461 & 1.457 & -0.733 & -0.735 \\ 
     5 & 0.004 & 0.935 & 0.954 & 1.475 & 1.472 & -0.705 & -0.706 \\ 
     6 & 0.001 & 0.934 & 0.956 & 1.258 & 1.256 & -0.656 & -0.646 \\ 
     7 & 0.002 & 0.930 & 0.948 & 1.265 & 1.269 & -0.591 & -0.585 \\ 
     8 & 0.004 & 0.930 & 0.954 & 1.181 & 1.178 & -0.495 & -0.516 \\ 
     9 & 0.003 & 0.931 & 0.956 & 1.187 & 1.187 & -0.442 & -0.459 \\ 
    10 & 0.003 & 0.952 & 0.943 & 1.141 & 1.140 & -0.403 & -0.423 \\ 
    15 & 0.002 & 0.942 & 0.965 & 1.100 & 1.100 & -0.276 & -0.289 \\ 
    25 & 0.005 & 0.934 & 0.955 & 1.057 & 1.057 & -0.143 & -0.159 \\ 
    50 & -0.002 & 0.929 & 0.959 & 1.027 & 1.026 & -0.053 & -0.075 \\ 
   100 & 0.000 & 0.920 & 0.946 & 1.013 & 1.013 & 0.022 & -0.017 \\ 
   \hline
\end{tabular}
\end{table}

%% --------------------------------
\subsection{Convergence diagnostics}
It is apparent that there is a relation between convergence diagnostics $\widehat{R}$ and $AC$, and the number of iterations per simulation condition($T$). Generally speaking, we see a trend towards convergence as $T$ increases with each simulation condition ($\widehat{R}$-values approach one, and $AC$-values approach zero). The convergence diagnostics show more or less equivalent trends for chain means versus chain variances. We, therefore, only discuss the $\widehat{R}$ and $AC$-values of chain means. %convergence diagnostics for the chain means. %do not differentiate in the interpretation of chain means .

Figure \ref{fig:conv}A shows that $\widehat{R}$-values generally decrease with increasing imputation chain lengths. The decline stabilizes somewhere between the simulation conditions $T=30$ and $T=50$. The downward trend is most pronounced for $T=3$, and between $T = 5$ and $T = 10$. In the intervening conditions ($3 \leq T \leq 5$), however, we observe a steep increase in $\widehat{R}$-values. This increase implies that convergence cannot be diagnosed at $T=3$, as would be the case based on the conventional threshold $\widehat{R} < 1.2$. \textbf{[This may be spurious??]} According to the widely used threshold $\widehat{R} < 1.1$, convergence can be diagnosed for conditions where $T>9$. If we use the recently recommended threshold $\widehat{R} < 1.01$, we would conclude that we cannot diagnose convergence in any simulation condition. \textbf{[Add something about $\widehat{R} <1$?]}

@@

The $AC$-values displayed in Figure \ref{fig:conv}B are almost uniformly increasing as a function of $T$. The only $AC$-value that deviates from this observation is for $T=2$. The lowest $AC$-value is obtained for $T=3$. At $T=5$, the $AC$-value reaches the level observed at $T=2$. Between $T=10$ and $T=30$ the increase plateaus off, and $AC$-values are stable around $T=70$. $AC$ is then indifferentiable from zero, indicating stationarity. %Simulation conditions where $T>5$ have $AC$-values greater than $T=2$. %We see an initial decrease between $T=2$ and $T=3$.
According to this diagnostic, none of the simulation conditions show signs of non-convergence, since we only observe negative or zero $AC$-values.

% All observed autocorrelations  are negative or close to zero. There are some differences between variables, and the ACs of chain variances seem lower than the ACs of chain means. The general trend in figure \ref{fig:conv} is that ACs decrease steeply as the number of iterations increases.  \textbf{The lowest AC observed is %-0.735 
% at four iterations. Simulation conditions with a higher number of iterations show a slow increase towards zero. autocorrelations do not plateau completely within the range of simulation conditions considered in this study. It seems that ACs might become positive if more iterations would be allowed. The average absolute maximum ACs are negative up to 77 iterations for chain means, and for chain variances all iterations considered here have negative ACs.}

@@

Taken together, we see that convergence may be diagnosed in simulation conditions where $T>10$.

%% --------------------------------
\subsection{Simulation diagnostics}
We use average bias, average confidence interval width, and coverage rate as performance measures to evaluate $\widehat{R}$ and $AC$. We make the general observation that the simulation diagnostics behave as theorized for most simulation conditions (bias around zero, coverage rate around 95\%).

>>>>

%The potential scale reduction factor $\widehat{R}$ is obtained as:
We can interpret $\widehat{R}$ as potential scale reduction factor since it indicates by how much the variance of $\theta$ could be shrunken down if an infinite number of iterations per chain would be run \citep{gelm92}. This interpretation assumes that chains are `over-dispersed' at $t=1$, and reach convergence as $T \to \infty$. Over-dispersion implies that the initial values of the chains are `far away' from the target distribution and each other. When all chains sample independent of their initial values, the mixing component of convergence is satisfied, and $\widehat{R}$-values will be close to one. High $\widehat{R}$-values thus indicate non-convergence. %The iteration at which chains are independent of their initial values is the point where mixing occurs. This is diagnosed with $\widehat{R}$-values close to one. %If there is no added value 
%High $\widehat{R}$-values thus indicate slow mixing of chains, which is a sign of non-convergence. %, e.g., because chains may depend on their initial value. %are of dependence across chains on initial values. Once there is no dependence on initial values anymore, there is no added value of continuing iterations. \textbf{[MOGELIJK OVERBODIGE ZIN HIER]} The measure then diagnoses algorithmic convergence ($\widehat{R} = 1$). 
%If there is no added value compared to infinitely many iterations, the chains have converged. 
The conventionally acceptable threshold for convergence was $\widehat{R} < 1.2$ \citep{gelm92}. More recently, \cite{veht19} proposed a more stringent threshold of $\widehat{R} < 1.01$.

@@

We can interpret %the sign and magnitude of ... Any non-zero $AC$-value ($-1 \leq AC \leq 1$) . $AC = 0$ indicates independence between $\theta$-values.
$AC$-values as a measure of stationarity. If $AC$-values are close to zero, there is no dependence between subsequent samples within imputation chains. Positive $AC$-values, however, indicate recurrence. If $\theta$-values of subsequent iterations are similar, trending may occur. Negative $AC$-values show no threat to the stationarity component of convergence. % (i.e., similarity between $\theta$-values of subsequent iterations), which may lead to trending. Negative $AC$-values occur when the $\theta$-values of subsequent iterations diverge from one-another. Dependence in this sense does not threaten the stationarity component of convergence. %that we aim to evaluate with $AC$. 
On the contrary even---negative $AC$-values indicate that $\theta$-values of subsequent iterations diverge from one-another, which may increase the variance of $\theta$ and speed up convergence.  %have also indicate dependence, but do not threaten the stationarity component of convergence \textbf{[REMOVE INTERPRETATION HERE?]}. Instead of recurrence, there would be a `seesaw' effect within chains: subsequent $\theta$-values would deviate in opposite directions from the chain average $\theta^{(\cdot m)})$, increasing the variance of $\theta$. 
As convergence diagnostic, the interest is therefore in positive $AC$-values.%\footnotemark 
\footnote{Moreover, the magnitude of $AC$-values may be evaluated statistically, but that is outside of this note's scope.}

In short, convergence is reached when there is no dependency between subsequent iterations of imputation chains ($AC = 0$), and chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} = 1$).

@@

This study evaluates whether $\widehat{R}$ and $AC$ could diagnose convergence of multiple imputation algorithms. %There is, however, no baseline measure available to evaluate their performance against.
%\footnote{The current practice of visually inspecting imputation chains for signs of non-mixing or non-stationarity can only diagnose severely pathological cases of non-convergence.} 
We assess the performance of the two convergence diagnostics against the recommended evaluation criteria for MI methods \citep[i.e., average bias, average confidence interval width, and empirical coverage rate across simulations;][\S~2.5.2]{buur18}. %The performance diagnostics comprise of .% %"We expect that in accordance with Neyman (1934), at least 95 percent of the confidence intervals should contain the true population value, although some room for simulation error should be taken into account to counteract the finite nature of our simulations"\cite{schou18}.
%\footnote{We could also look at distributional characteristics, and plausibility of imputed values, see \cite{vinknd} (n.d.). For now, this is outside of the scope of this study.} % Or use: Then evaluate the regular diagnostics of MI simulations, see Vink, n.d., and  convergence. Diagnostics include absolute bias of the estimated regression coefficient, confidence interval width, and empirical coverage rate across simulations. Convergence is evaluated with $\widehat{R}$, or potential scale reduction factor.

%\textbf{[HIER NOG ZOOI]} We hypothesize that $\widehat{R}$ may not be an appropriate diagnostic, because it assumes over-dispersed initial values. In \pkg{mice}, initial values of the algorithm are chosen randomly from the observed data. Therefore, we cannot be certain that the initial values are over-dispersed.
%The initial parameter value is noodzakelijk. Daarom is het lastig om in MI that starten vanuit overdisp state. 
% Starting values of MICE chains: "For each j, fill in starting imputations $Y^0_j$ by random draws from $Y^{obs}_j$." \cite[\S~4.5.2]{buur18} before for loops over variables and imputations, see \url{https://stefvanbuuren.name/fimd/sec-FCS.html#def:mice}
%Without over-dispersed initial states, $\widehat{R}$ may falsely diagnose convergence \citep{broo98}. %would not be sensitive enough to flag non-convergence of MI algorithms: "if over-dispersion does not hold, $\sigma_+^2$ can be too low, which can lead to falsely diagnosing convergence." \cite[p~437]{broo98}. 
%This suggests that $\widehat{R}$ would not be sensitive enough to flag non-convergence of MI algorithms. However, an empirical finding by \cite{lace07} shows that $\widehat{R}$ can indicate non-convergence at unconventionally long chain lengths. While the default number of iterations in \pkg{mice} is five, $\widehat{R} > 1.1$ was observed after 50 iterations, and $\widehat{R} > 1.01$ up-to 150 iterations. This suggests that $\widehat{R}$ would over-estimate non-convergence of MI algorithms.

Based on an empirical finding \citep{lace07}, we hypothesize that $\widehat{R}$ will over-estimate non-convergence of MI algorithms. The threshold of $\widehat{R} < 1.01$ will then be too stringent for diagnosing convergence. This over-estimation may, however, be diminished because $\widehat{R}$ can falsely diagnose convergence if initial values of the algorithm are not appropriately over-dispersed \cite[p~437]{broo98}. In \pkg{mice}, initial values are chosen randomly from the observed data. Therefore, we cannot be certain that the initial values are over-dispersed. We expect this to have little effect on the hypothesized performance of $\widehat{R}$. No hypothesis was formulated about the performance of $AC$ as convergence diagnostic.
%

@@

%To investigate the convergence properties of the multiple imputation algorithm in \pkg{mice}, w
We use model-based simulation in \proglang{R} to perform multiple imputation under 100 simulation conditions. In each condition, the MI algorithm comprises of a different imputation chain length ($T = 1, 2, \dots, 100$). The number of simulation runs per condition is 1000. For each simulation condition (i.e., number of iterations) and each repetition (i.e., simulation run) we compute convergence diagnostics ($\widehat{R}$ and autocorrelation) and simulation diagnostics (bias, confidence interval width and coverage). These diagnostics are aggregated across repetitions to obtain their average value per simulation condition.
%We compute $\widehat{R}$ and autocorrelation (`convergence diagnostics') and performance measures (`simulation diagnostics') for each condition. %All simulation conditions are replicated 1000 times. %For each of the 100 conditions 1000 simulation runs are performed. for each of the 100 simulation conditions. \textbf{Estimands are convergence diagnostics. Performance measures are simulation diagnostics.  where for one through one hundred the Rhat and ... are calculated.} 
The simulation set-up is summarized in the pseudo-code below.\footnote{The complete \proglang{R} script of the simulation study is available from \href{https://github.com/gerkovink/shinyMice/simulation}{github.com/gerkovink/ShinyMICE/simulation}.}

@@

\subsection{Amputation}
The complete data is `amputed' once for each simulation repetition. That is, the function \fct{mice::ampute} is used to impose a missingness mechanism upon the data. The missingness is univariate, and the probability to be missing is the same for all four variables, namely 20\% (\texttt{prop = 0.8, mech = "MCAR"}). This leaves 20\% of the rows completely observed. %The resulting amputed data is equal for all simulation conditions in the same repetition. %"Furthermore, we generate three kinds of missingness proportions: 0.1, 0.5, and 0.9. Note that these values indicate the sampled proportion of incomplete cases in Y" \cite{schou18}.

\subsection{Imputation}
Missing datapoints are imputed with the function \fct{mice::mice}. All MI procedures are performed with Bayesian linear regression imputation (\texttt{method = "norm"}), and five imputation chains (\texttt{m = 5}). The number of iterations varies between simulation conditions (\texttt{maxit = 1, 2, \dots, 100}). 
% Missingness mechanism: "With MCAR missingness mechanisms, the probability to be missing is the same for all cases. This is a necessary simulation condition for evaluating the performance of imputation procedures. If an imputation method is not able to solve the problem (i.e. yield valid inference) under MCAR, the statistical properties of the procedure are not sound." (Vink, n.d., p. 4)
% "combine the m completed data inferences into a single inference following Rubin's (1987) rules (pp. 76-77)" \cite{schou18}.

\subsection{Convergence Diagnostics}
% Add ``post-processing steps (e.g. aggregation computations, summary statistics, regressions on the simulation results) used to transform simulation outputs to reported results''. %, see \url{http://jsterman.scripts.mit.edu/docs/Rahmandad-Sterman%20Simulation%20Reporting%20Standards3.30%20update.pdf}.
From the imputation step, we extract chain means and chain variances across iterations ($\theta$ at $t = 1,2,\dots,T$). %not only the imputed values of each MI procedure (where $t = T$), but also information from the intermediate iterations ($t = 1, 2, \dots, T$). The quantities of interest ($\theta$-values) are chain means and chain variances. 
We compute convergence diagnostics separately for both $\theta$s, and for each of the four variables in the dataset. We report the maximum (absolute) value as the `worst linear predictor' of convergence. $\widehat{R}$ is computed by implementing Vehtari et al.'s (\citeyear{veht19}) recommendations. $AC$ is computed with function \fct{stats::acf}.
%%%%%%%%%%%%%%%%
%From each imputation chain, we extract imputed values, chain means and chain variances ($\theta$-values). $\theta$-values are information from the intermediate iterations ($t = 1, 2, \dots, T$). The quantities of interest ($\theta$-values) are . We compute convergence diagnostics separately for each of the four variables in the dataset. \textbf{We report the maximum (absolute) value as the `worst linear predictor' of convergence [Rephrase or footnote??]}. $\widehat{R}$ is computed by implementing Vehtari et al.'s \citeyear{veht19} recommendations. $AC$ is computed with \pkg{stats} function \fct{acf}.
%%%%%%%%%%%%%
% 
% 
% extract chain means and chain variances per iteration $t$. From the MI algorithm Convergence diagnostics are computed for chain means and chain variances.   
% $\widehat{R}$ is computed within imputation chains: . The maximum value across variables within the same simulation is reported. 
% 
% autocorrelation (AC) is computed within imputation chains,  The AC of the variable with the highest absolute AC value is reported per simulation.

@@

To estimate the quantity of scientific interest, $Q$, we perform multiple linear regression on each completed dataset with the function \fct{stats::lm}. We obtain an estimated regression coefficient per imputation, which are pooled into a single estimate, $\bar{Q}$. We use the function \fct{mice::pool} to get variance estimates according to Rubin's (\citeyear{rubin87}) rules, and subsequently implement finite population pooling conform \cite{vink14}.

@@

We compute bias as the difference between $\bar{Q}$ and $Q$. Confidence interval width (CIW) is defined as the difference between the lower and upper bound of the 95\% confidence interval (CI95\%) around $\bar{Q}$. We compute the CI95\% bounds as

$$\bar{Q} \pm t_{(m-1)} \times SE_{\bar{Q}},$$

where $t_{(m-1)}$ is the quantile of a $t$-distribution with $m-1$ degrees of freedom, and $SE_{\bar{Q}}$ is the square root of the pooled variance estimate. From bias and CIW, we calculate empirical coverage rates. Coverage rate is the proportion of simulations in which $Q$ is between the bounds of the CI95\% around $\bar{Q}$.

@@

% \begin{figure}[h]
%   \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
%         \includegraphics{Figures/conv_diag_SEs.pdf}
%   }
%   \caption{Convergence diagnostics over 1000 MCMC simulations.}
%     \label{fig:conv2}
% \end{figure}

\begin{figure}[h]
  \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
        \includegraphics{Figures/simulation_diag.pdf}
  }
  \caption{Simulation diagnostics over 1000 MCMC simulations. The dashed line depicts the average trend across simulation conditions (Loess line).}
    \label{fig:sim}
\end{figure}

% \begin{figure}[h]
%   \resizebox{\textwidth}{!}{ %notice the \resizebox{} command
%         \includegraphics{Figures/sim_diag_SEs.pdf}
%   }
%   \caption{Simulation diagnostics over 1000 MCMC simulations.}
%     \label{fig:sim2}
% \end{figure}

latex table generated in R 3.6.1 by xtable 1.8-4 package
Wed Dec 04 18:16:11 2019
\begin{table}[ht]
\centering
\caption{Simulation and convergence diagnostics over 1000 MCMC simulations. \footnotesize{No convergence diagnostics are reported for the condition where T=1, since these relative measures cannot be computed for a single value.}}
\label{tab:results}
\begin{tabular}{lrrrrrrr}
  \hline
T & Bias & CI width & Cov. rate & $\widehat{R}_{mean}$ & $\widehat{R}_{var}$ & $AC_{mean}$ & $AC_{var}$ \\
  \hline
   1 & -0.137 & 0.954 & 0.932 & NA & NA & NA & NA \\
     2 & -0.006 & 0.932 & 0.953 & 1.650 & 1.632 & -0.500 & -0.500 \\
     3 & 0.002 & 0.929 & 0.944 & 1.314 & 1.306 & -0.660 & -0.659 \\
     4 & 0.003 & 0.933 & 0.957 & 1.461 & 1.457 & -0.733 & -0.735 \\
     5 & 0.004 & 0.935 & 0.954 & 1.475 & 1.472 & -0.705 & -0.706 \\
     6 & 0.001 & 0.934 & 0.956 & 1.258 & 1.256 & -0.656 & -0.646 \\
     7 & 0.002 & 0.930 & 0.948 & 1.265 & 1.269 & -0.591 & -0.585 \\
     8 & 0.004 & 0.930 & 0.954 & 1.181 & 1.178 & -0.495 & -0.516 \\
     9 & 0.003 & 0.931 & 0.956 & 1.187 & 1.187 & -0.442 & -0.459 \\
    10 & 0.003 & 0.952 & 0.943 & 1.141 & 1.140 & -0.403 & -0.423 \\
    15 & 0.002 & 0.942 & 0.965 & 1.100 & 1.100 & -0.276 & -0.289 \\
    25 & 0.005 & 0.934 & 0.955 & 1.057 & 1.057 & -0.143 & -0.159 \\
    50 & -0.002 & 0.929 & 0.959 & 1.027 & 1.026 & -0.053 & -0.075 \\
   100 & 0.000 & 0.920 & 0.946 & 1.013 & 1.013 & 0.022 & -0.017 \\
   \hline
\end{tabular}
\end{table}

%% --------------------------------
\subsection{Convergence diagnostics}
It is apparent that there is a relation between the number of iterations per simulation condition ($T$) and the convergence diagnostics. Generally speaking, conditions with longer imputation chains (higher $T$) coincide with less signs of non-convergence ($\widehat{R}$-values approach one, and $AC$-values approach zero). We see more or less equivalent trends for chain means versus chain variances. We, therefore, only discuss the $\widehat{R}$ and $AC$-values of chain means. %convergence diagnostics for the chain means. %do not differentiate in the interpretation of chain means .

Figure \ref{fig:conv}A shows that $\widehat{R}$-values generally decrease with increasing imputation chain lengths. The decline stabilizes somewhere between the simulation conditions $T=30$ and $T=50$. The downward trend is most pronounced for $T=3$, and between $T = 5$ and $T = 10$. In the intervening conditions ($3 \leq T \leq 5$), however, we observe a steep increase in $\widehat{R}$-values. This increase implies that non-convergence is under-estimated, and convergence should not be diagnosed at this point. Based on the conventional threshold $\widehat{R} < 1.2$, we would falsely diagnose convergence at $T=3$. According to the widely used threshold $\widehat{R} < 1.1$, convergence would be diagnosed for conditions where $T>9$. If we use the recently recommended threshold $\widehat{R} < 1.01$, we would conclude that convergence is reached in none of the simulation conditions. %\textbf{[Add something about $\widehat{R} <1$?]}

@@

The $AC$-values displayed in Figure \ref{fig:conv}B are almost uniformly increasing as a function of $T$. The only $AC$-value that deviates from this observation is for $T=2$. The lowest $AC$-value is obtained for $T=3$. %At $T=5$, the $AC$-value reaches the level observed at $T=2$. 
The gradual increase plateaus between $T=10$ and $T=30$. $AC$-values in conditions where $T>70$ are indifferentiable from zero, indicating stationarity. %Simulation conditions where $T>5$ have $AC$-values greater than $T=2$. %We see an initial decrease between $T=2$ and $T=3$.
According to this diagnostic, none of the simulation conditions show signs of non-convergence, since we only observe negative or zero $AC$-values. The negative $AC$-values, however, indicate increasing divergence between subsequent samples in conditions where $T<4$. It may not be wise to terminating the algorithm at that point, but after `recovering' from this dip in $AC$-values. This would imply $T=6$ as the minimal number of iterations. 
% All observed autocorrelations  are negative or close to zero. There are some differences between variables, and the ACs of chain variances seem lower than the ACs of chain means. The general trend in figure \ref{fig:conv} is that ACs decrease steeply as the number of iterations increases.  \textbf{The lowest AC observed is %-0.735 
% at four iterations. Simulation conditions with a higher number of iterations show a slow increase towards zero. autocorrelations do not plateau completely within the range of simulation conditions considered in this study. It seems that ACs might become positive if more iterations would be allowed. The average absolute maximum ACs are negative up to 77 iterations for chain means, and for chain variances all iterations considered here have negative ACs.}

@@

Taken together, we see that $T>3$ is the minimal requirement to diagnose convergence ($\widehat{R} < 1.2; AC \leq 0$). This threshold is, however, not sufficient, since we overlook the increase in $\widehat{R}$-values up-to conditions where $T>5$, and the convergence diagnostics only reach stability at $T>20$. %This implies that the current

%% --------------------------------
\subsection{Simulation diagnostics}
We use average bias, average confidence interval width, and coverage rate as performance measures to evaluate $\widehat{R}$ and $AC$. We make the general observation that the simulation diagnostics behave as theorized for most simulation conditions (bias around zero, stable confidence interval widths, nominal coverage rate at 95\%).
